PaperMelanoma

### 2022-06-14 Code/PaperMelanoma.Rmd
Final checks

### Exclude patients
Exclude when Exclusion = 2: based on duplicates, urogenital melanoma, etc.
Leave the patients included that have EXCLUSIE = 1 and Dbstatus = 8,
even though diagnosis and operation date are very close.
```{r, eval=TRUE}
sel.data <- original.data[original.data$EXCLUSIE!="Exclusion", selected.vars]
cat("Number of patients excluded:", nrow(original.data)-nrow(sel.data), "\n")
```
Omit patient A0523 - no last follow-up date
```{r, eval=TRUE}
sel.data <- sel.data[sel.data$DatabaseNR!="A0523               ",]
```

### Recode
Set Rdamcrit of negative SN patients equal to 0. 11 patients with positive SN have SN tumour burden of exactly 1
```{r, eval=TRUE}
sum(sel.data$Rdamcrit==1&sel.data$SNstatus=="Positive"&!is.na(sel.data$Rdamcrit))
sel.data$Rdamcrit[sel.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
```

Merge single cells with subcap and set Dewar for negative patients to subcap
```{r, eval=TRUE}
sel.data$Dewar[sel.data$Dewar=="single cells"] <- "subcap"
sel.data$Dewar[sel.data$SNstatus=="Negative"] <- "subcap"
```

Merge location other with trunk
```{r, eval=TRUE}
sel.data$Loc_CAT[sel.data$Loc_CAT=="other"] <- "trunk"
```

@@

Melanoma specific mortality (MSM)
```{r, eval=TRUE}
sel.data$MSM<-1*(sel.data$Status=="DOD")
```

Define multiple fields
```{r, eval=TRUE}
sel.data$multiple.fields <- ifelse(!is.na(sel.data$nr_SNs_Field_2)&sel.data$nr_SNs_Field_2>0,1,0)
```

@@

Use age at SN instead of age at diagnosis
```{r, eval=TRUE}
sel.data$Age.SN <- floor((sel.data$SNdate-sel.data$DOB)/(24*60*60)/365.25)
```

Define time since diagnosis in years (follow-up time)
```{r, eval=TRUE}
sel.data$FU.time <- (sel.data$Last_FU-sel.data$SNdate)/(24*60*60)/365.25
```

@@

Set Mitosis to Unknown if missing
```{r, eval=TRUE}
sel.data$Mitosis[is.na(sel.data$Mitosis)] <- "Unknown" # mitosis unknown for 80% of patients
```

Replace "Unknown" by NA
```{r, eval=TRUE}
sel.data$Ulceration[sel.data$Ulceration=="unknown"]<-NA
sel.data$Loc_CAT[sel.data$Loc_CAT=="unknown"]<-NA
sel.data$Histology[sel.data$Histology=="unknown"]<-NA
sel.data$Dewar[sel.data$Dewar=="unknown"] <- NA
sel.data$Mitosis[sel.data$Mitosis=="Unknown"] <- NA
```

Delete unused levels
```{r, eval=TRUE}
sel.data <- droplevels(sel.data)
```

@@

### Restrict survival at 5 years
```{r, eval=TRUE}
S.Rec[S.Rec[,1]>5, 1] <- 5 # set recurrence time to 5
S.Rec[S.Rec[,1]>5, 2] <- 0 # set censoring
S.MSM[S.MSM[,1]>5, 1] <- 5 # set follow-up time to 5
S.MSM[S.MSM[,1]>5, 2] <- 0 # set censoring
```

@@

# Calibrated final model for melanoma specific mortality
Fit disease specific mortality to the linear predictor of recurrence
```{r, eval=TRUE}
lp.BS<-f.mi.BS.Rec$linear
f.mi.BS.MSM<-rms::cph(S.MSM~lp.BS,y=TRUE,x=TRUE)
utils::write.table(paste0(sprintf("%.2f", stats::coef(f.mi.BS.MSM)), " [",
            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM))[1]), "; ",
            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM))[2]), "]"),
            file='Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt', row.names=FALSE, col.names=FALSE)
MSM.cal.fact <- stats::coef(f.mi.BS.MSM)
stats::coef(f.mi.BS.MSM)
stats::confint(f.mi.BS.MSM)
```

@@

# Plot calibration of all centers combined and all centers seperate 
for recurrence and melanoma specific mortality refit 
write results to Excel
```{r, eval=TRUE}
horizon <- 5
for (model in list('Rec', 'MSM.refit')){
  if (model == "Rec"){
    # plot predict
    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Rec.png"))
    plot.predict <- plot(Predict(f.mi.full.Rec))
    show(plot.predict)
    grDevices::dev.off()

@@

# Baseline hazard
rc.Rec <- (1/attributes(nom.0)$info$sc)
int.Rec <- attributes(nom.0)$info$Intercept
f.Rec.basehaz <- basehaz(f.mi.BS.Rec,TRUE)
h0.Rec <- f.Rec.basehaz$hazard[f.Rec.basehaz$time==max(f.Rec.basehaz$time[f.Rec.basehaz$time<=horizon])]
lp.range <- int.Rec+rc.Rec*(0:10)*2
par(mar = rep(0, 4))
nom <- nomogram(f.mi.BS.Rec, maxscale=10,
              Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class,
              lp=TRUE, fun.at=fun.event(lp=lp.range, h0=h0.Rec))
attributes(nom)$names[attributes(nom)$names=="Age.SN"]<-"Age"
attributes(nom)$names[attributes(nom)$names=="Loc_CAT"]<-"Location"
attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Tumor burden"
attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Positive)"]<-"Breslow + positive SN"
attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow + negative SN"
plot(nom, total.sep.page=TRUE, col.grid=gray(c(0.8, 0.95)))
nom.0
nom
```

# Risk score distribution
```{r, eval=TRUE}
rc.MSM <- f.mi.BS.MSM$coef[1]*rc.Rec
int.MSM <- f.mi.BS.MSM$coef[1]*int.Rec

max.range <- 16
points.range <- 0:max.range
lp.points.Rec <- int.Rec+rc.Rec*points.range
lp.points.MSM <- int.MSM+rc.MSM*points.range

f.MSM.basehaz <- basehaz(f.mi.BS.MSM,TRUE)
h0.MSM <- f.MSM.basehaz$hazard[f.MSM.basehaz$time==max(f.MSM.basehaz$time[f.MSM.basehaz$time<=horizon])]
lp.MSM <- f.mi.BS.MSM$linear

Breslow.trunc <- last.imputed.data.set$Breslow
Breslow.trunc[last.imputed.data.set$Breslow<0.1] <- .1
Breslow.trunc[last.imputed.data.set$Breslow>7] <- 7
lp.score.Rec <- predict(f.mi.BS.Rec,newdata=data.frame(SNstatus=last.imputed.data.set$SNstatus,
                                                Age.SN=last.imputed.data.set$Age.SN,
                                                Ulceration=last.imputed.data.set$Ulceration,
                                                Loc_CAT=last.imputed.data.set$Loc_CAT,
                                                Breslow=Breslow.trunc,
                                                Rdamcrit=last.imputed.data.set$Rdamcrit))
mean(lp.score.Rec)
score.Rec <- round((lp.score.Rec-int.Rec)/rc.Rec,0)
max.breaks <- max.range+1
h.Rec <- hist(score.Rec,plot=FALSE,breaks=0:max.breaks,right=FALSE)

@@

```{r, eval=TRUE, echo=FALSE}
par(mar = c(5,5,2,5))
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col='white')
abline(h=(0:30)*1,col="light grey",lty=2)
axis(side = 4)
mtext(side = 4, line = 3, 'Risk score distribution (%)')
par(new = TRUE)
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")

par(new = TRUE)
plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg='white')
```

@@

# Test if model is correct
### Check calibration
```{r, eval=TRUE}
# Slope and intercept of recurrence and MSM
int.Rec
# check that int is same as this:
log(20)*f.mi.BS.Rec$coef["Age.SN"]+log(0.1)*f.mi.BS.Rec$coef["Breslow"]+log(0.01)*f.mi.BS.Rec$coef["Rdamcrit"]-f.mi.BS.Rec$center
```

### Check mean event rates of risk distribution
```{r, eval=TRUE}
# Recurrence
sum(h.Rec$density*fun.event(lp=lp.points.Rec, h0=h0.Rec))
# equal to the 5-year survival probability, kaplanmeier op 5 jaar, survfit op 5 jaar
SF<-survfit(S.Rec~1,data=sel.data)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))

# MSM
sum(h.Rec$density*fun.event(lp=lp.points.MSM, h0=h0.MSM))
SF<-survfit(S.MSM~1,data=sel.data)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
mean(fun.event(lp=lp.MSM, h0=h0.MSM))
```

@@

# score
round((lp.score.Rec.try-int.Rec)/rc.Rec,0)

# manual calculation
coef.Rec <- stats::coef(f.mi.BS.Rec)
lp.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)+
              coef.Rec["Age.SN"]*log(data.to.be.imputed[patient.nr,"Age.SN"])+
              coef.Rec["Ulceration=Yes"]*(as.numeric(data.to.be.imputed[patient.nr,"Ulceration"])-1)+
              coef.Rec["Loc_CAT=leg"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="leg")+
              coef.Rec["Loc_CAT=trunk"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="trunk")+
              coef.Rec["Loc_CAT=head & neck"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="head & neck")+
              coef.Rec["Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)+
              coef.Rec["Rdamcrit"]*log(data.to.be.imputed[patient.nr,"Rdamcrit"])+
              coef.Rec["SNstatus=Positive * Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)
center.Rec <- f.mi.BS.Rec$center

# lp
lp.score.Rec.true
lp.score.Rec.try
lp.manual-center.Rec

# probability of recurrence
fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
1-exp(-h0.Rec*exp(lp.manual-center.Rec))

# check MSM
lp.score.MSM.true <- predict(f.mi.BS.MSM)[patient.nr]
lp.score.MSM.true
lp.MSM.try <- MSM.cal.fact*(lp.manual-center.Rec)
lp.MSM.try
fun.event(lp=lp.MSM.try, h0=h0.MSM)
fun.event(lp=lp.MSM.try, h0=h0.MSM)
1-exp(-h0.MSM*exp(MSM.cal.fact*(lp.manual-center.Rec)))
```

### Test with old tool predictions recurrence for negative SN status patients
```{r, eval=TRUE}
rscript.negative <- function(breslow, ulceration, location) {
  S0 <- 0.85837329
  lp <- 0.79351989*log(breslow) +
    0.61912477*I(ulceration == 1) +
    0.29594843*I(location == 1) +
    0.42802046*I(location == 2) +
    0.86053044*I(location == 3) - 0.98860941
  S <- 100 - 100*S0^exp(lp)
  return(S)
}
old.negativeSN.tool <- rscript.negative(breslow=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"],
        ulceration=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Ulceration"])-1,
        location=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Loc_CAT"])-1)
lp.negative <- predict(f.mi.BS.Rec, newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative",], type="lp")
new.negativeSN.tool <- 1-exp(-h0.Rec*exp(lp.negative))
plot(x=fun.event(lp=new.negativeSN.tool, h0=h0.Rec)*100, y=old.negativeSN.tool,
     xlim=c(0, 100), ylim=c(0, 100),
     xlab="New predictions", ylab="Old predictions",
     main="Predicted recurrence of patients with negative SN status")
abline(a=0, b=1, col="red")
```

### Test with old tool predictions recurrence for positive SN status patients
```{r, eval=TRUE}
rscript.positive <- function(age, ulceration, tumor_burden, breslow) {
  S0 <- 0.51762362
  lp <- 0.56109945 * log(age) +
    0.23383770 * log(tumor_burden) +
    0.35489235 * log(breslow) +
    0.36562098 * I(ulceration == 1) - 2.709225
  S <- 100 - 100 * S0^exp(lp)
  return(S)
}
old.positiveSN.tool <- rscript.positive(age=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Age.SN"],
                               ulceration=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Ulceration"])-1,
                               breslow=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"],
                               tumor_burden=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Rdamcrit"]))
lp.positive <- predict(f.mi.BS.Rec, newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive",], type="lp")
new.positiveSN.tool <- 1-exp(-h0.Rec*exp(lp.positive))
plot(x=fun.event(lp=new.positiveSN.tool, h0=h0.Rec)*100, y=old.positiveSN.tool,
     xlim=c(0, 100), ylim=c(0, 100),
     xlab="New predictions", ylab="Old predictions",
     main="Predicted recurrence of patients with positive SN status")
abline(a=0, b=1, col="red")
```

>>>>

### Exclude patients
Exclude when Exclusion = 2: based on duplicates, urogenital melanoma, etc. Note:
Leave the patients included that have EXCLUSIE = 1 and Dbstatus = 8 (Dbstatus=0: 
negative, Dbstatus=1: positive), even though diagnosis and operation date are 
very close.
```{r, eval=TRUE}
sel.data <- original.data[original.data$EXCLUSIE!="Exclusion", selected.vars]
cat("Number of patients excluded:", nrow(original.data)-nrow(sel.data), "\n")
```
Omit patient A0523 - no last follow-up date
```{r, eval=TRUE}
sel.data[which(is.na(sel.data$Last_FU)), "DatabaseNR"]
sel.data <- sel.data[sel.data$DatabaseNR!="A0523               ",]
```

### Recode
#### Use age at SN instead of age at diagnosis
```{r, eval=TRUE}
sel.data$Age.SN <- floor((sel.data$SNdate-sel.data$DOB)/(24*60*60)/365.25)
```

#### Location
Merge location other with trunk
```{r, eval=TRUE}
sel.data$Loc_CAT[sel.data$Loc_CAT=="other"] <- "trunk"
```

@@

#### Define multiple fields
```{r, eval=TRUE}
sel.data$multiple.fields <- ifelse(!is.na(sel.data$nr_SNs_Field_2)&sel.data$nr_SNs_Field_2>0,1,0)
```

#### Number of positive and negative nodes
Check: there are no negative patients that have positive nodes
```{r, eval=TRUE}
sum(!is.na(sel.data$Tot_SNs_pos)&sel.data$Tot_SNs_pos>0&sel.data$SNstatus=="Negative")
```

@@

467 positive SN patients have negative nodes
```{r, eval=TRUE}
sum(!is.na(sel.data$Tot_SNs_neg)&sel.data$Tot_SNs_neg>0&sel.data$SNstatus=="Positive")
```

#### SN tumour burden
11 patients with positive SN have SN tumour burden of exactly 1
```{r, eval=TRUE}
sum(sel.data$Rdamcrit==1&sel.data$SNstatus=="Positive"&!is.na(sel.data$Rdamcrit))
```

Set Rdamcrit of negative SN patients equal to 0
```{r, eval=TRUE}
sel.data$Rdamcrit[sel.data$SNstatus=="Negative"] <- 1 # since log(1) = 0
```

#### Location of metastasis
Set Dewar for negative patients to subcap
```{r, eval=TRUE}
sum(sel.data$SNstatus=="Positive"&sel.data$Dewar=="subcap"&!is.na(sel.data$Dewar))
sel.data$Dewar[sel.data$SNstatus=="Negative"] <- "subcap"
```

Merge single cells with subcap 
```{r, eval=TRUE}
sel.data$Dewar[sel.data$Dewar=="single cells"] <- "subcap"
```

#### Recurrence 
Define time since diagnosis in years (follow-up time)
```{r, eval=TRUE}
sel.data$FU.time <- (sel.data$Last_FU-sel.data$SNdate)/(24*60*60)/365.25
```

@@

#### Melanoma specific mortality (MSM)
```{r, eval=TRUE}
sel.data$MSM<-1*(sel.data$Status=="DOD")
```

#### Mitosis
Set Mitosis to Unknown if missing
```{r, eval=TRUE}
sel.data$Mitosis[is.na(sel.data$Mitosis)] <- "Unknown" # mitosis unknown for 80% of patients
```

#### Replace "Unknown" by NA
```{r, eval=TRUE}
sel.data$Ulceration[sel.data$Ulceration=="unknown"]<-NA
sel.data$Loc_CAT[sel.data$Loc_CAT=="unknown"]<-NA
sel.data$Histology[sel.data$Histology=="unknown"]<-NA
sel.data$Dewar[sel.data$Dewar=="unknown"] <- NA
sel.data$Mitosis[sel.data$Mitosis=="Unknown"] <- NA
```

#### Delete unused levels
```{r, eval=TRUE}
sel.data <- droplevels(sel.data)
```

@@

# Restrict survival at 5 years
```{r, eval=TRUE}
S.Rec[S.Rec[,1]>5, 1] <- 5 # set recurrence time to 5
S.Rec[S.Rec[,1]>5, 2] <- 0 # set censoring
S.MSM[S.MSM[,1]>5, 1] <- 5 # set follow-up time to 5
S.MSM[S.MSM[,1]>5, 2] <- 0 # set censoring
```

@@

# Calibrated final model for melanoma specific mortality
Fit disease specific mortality to the linear predictor of recurrence
```{r, eval=TRUE}
lp.BS<-f.mi.BS.Rec$linear
f.mi.BS.MSM<-rms::cph(S.MSM~lp.BS,y=TRUE,x=TRUE)
MSM.cal.fact <- stats::coef(f.mi.BS.MSM)
MSM.cal.fact
stats::confint(f.mi.BS.MSM)
utils::write.table(paste0(sprintf("%.2f", stats::coef(f.mi.BS.MSM)), " [",
            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM))[1]), "; ",
            sprintf("%.2f", as.numeric(stats::confint(f.mi.BS.MSM))[2]), "]"),
            file='Z:/Project Melanoom/PaperMelanoma/Results/calibration.MSS.txt', row.names=FALSE, col.names=FALSE)
```

@@

# Plot calibration of all centers combined and all centers seperate 
for recurrence and melanoma specific mortality refit 
```{r, eval=TRUE}
horizon <- 5
for (model in list('Rec', 'MSM.refit')){
  if (model == "Rec"){
    # plot predict
    png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/Figures/plot.predict.Rec.png"))
    plot.predict <- plot(Predict(f.mi.full.Rec))
    show(plot.predict)
    grDevices::dev.off()

@@

# baseline hazard of recurrence
f.Rec.basehaz <- basehaz(f.mi.BS.Rec,TRUE)
h0.Rec <- f.Rec.basehaz$hazard[f.Rec.basehaz$time==max(f.Rec.basehaz$time[f.Rec.basehaz$time<=horizon])]

# lp range Recurrence
rc.Rec <- (1/attributes(nom.0)$info$sc)
int.Rec <- attributes(nom.0)$info$Intercept
lp.range <- int.Rec+rc.Rec*(0:10)*2

# nomogram
par(mar = rep(0, 4))
nom <- nomogram(f.mi.BS.Rec, maxscale=10,
              Breslow=Breslow.class, Rdamcrit=Rdamcrit.class, Age.SN=Age.class,
              lp=TRUE, fun.at=fun.event(lp=lp.range, h0=h0.Rec))
attributes(nom)$names[attributes(nom)$names=="Age.SN"]<-"Age"
attributes(nom)$names[attributes(nom)$names=="Loc_CAT"]<-"Location"
attributes(nom)$names[attributes(nom)$names=="Rdamcrit"]<-"Tumor burden"
attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Positive)"]<-"Breslow + positive SN"
attributes(nom)$names[attributes(nom)$names=="Breslow (SNstatus=Negative)"]<-"Breslow + negative SN"
plot(nom, total.sep.page=TRUE, col.grid=gray(c(0.8, 0.95)))
nom.0
nom
```

# Risk score distribution
```{r, eval=TRUE}
# baseline hazard of MSM
f.MSM.basehaz <- basehaz(f.mi.BS.MSM,TRUE)
h0.MSM <- f.MSM.basehaz$hazard[f.MSM.basehaz$time==max(f.MSM.basehaz$time[f.MSM.basehaz$time<=horizon])]
lp.MSM <- f.mi.BS.MSM$linear

# lp range MSM
rc.MSM <- f.mi.BS.MSM$coef[1]*rc.Rec
int.MSM <- f.mi.BS.MSM$coef[1]*int.Rec
max.range <- 16
points.range <- 0:max.range
lp.points.Rec <- int.Rec+rc.Rec*points.range
lp.points.MSM <- int.MSM+rc.MSM*points.range

# Recurrence lp scores
Breslow.trunc <- last.imputed.data.set$Breslow
Breslow.trunc[last.imputed.data.set$Breslow<0.1] <- .1
Breslow.trunc[last.imputed.data.set$Breslow>7] <- 7
lp.score.Rec <- predict(f.mi.BS.Rec,newdata=data.frame(SNstatus=last.imputed.data.set$SNstatus,
                                                Age.SN=last.imputed.data.set$Age.SN,
                                                Ulceration=last.imputed.data.set$Ulceration,
                                                Loc_CAT=last.imputed.data.set$Loc_CAT,
                                                Breslow=Breslow.trunc,
                                                Rdamcrit=last.imputed.data.set$Rdamcrit))
mean(lp.score.Rec)
score.Rec <- round((lp.score.Rec-int.Rec)/rc.Rec,0)
max.breaks <- max.range+1
h.Rec <- hist(score.Rec,plot=FALSE,breaks=0:max.breaks,right=FALSE)

@@

```{r, eval=TRUE, echo=FALSE}
par(mar = c(5,5,2,5))
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col='white')
abline(h=(0:30)*1,col="light grey",lty=2)
axis(side = 4)
mtext(side = 4, line = 3, 'Risk score distribution (%)')
par(new = TRUE)
plot(h.plot,freq=FALSE,axes=FALSE, xlab=NA, xlim=x.lim, ylab=NA,ylim=y.lim.hist,main=NA,col="white")
par(new = TRUE)
plot(points.range,p.Rec*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=15, xlab="Risk score",ylab="Predicted 5-year risk (%)")
points(points.range,p.MSM*100,type="b",lwd=1,xlim=x.lim,ylim=y.lim,pch=16)
legend(x=0,y=100,legend=c("Recurrence","Melanoma-specific mortality"),pch=c(15,16),cex=.9,bg='white')
```

@@

# Test if model is correct
### Check calibration
Intercept of recurrence should be the same as when setting covariates to 'zero' values minus the center
```{r, eval=TRUE}
int.Rec
log(20)*f.mi.BS.Rec$coef["Age.SN"]+
  (log(0.1)-c.Breslow)*f.mi.BS.Rec$coef["Breslow"]+
  log(0.01)*f.mi.BS.Rec$coef["Rdamcrit"]-
  f.mi.BS.Rec$center
```

### Check mean event rates of risk distribution
#### Recurrence
```{r, eval=TRUE}
sum(h.Rec$density*fun.event(lp=lp.points.Rec, h0=h0.Rec))
# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
SF<-survfit(S.Rec~1,data=sel.data)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
# average probability of recurrence
mean(fun.event(lp=lp.score.Rec, h0=h0.Rec))
```

#### MSM
```{r, eval=TRUE}
sum(h.Rec$density*fun.event(lp=lp.points.MSM, h0=h0.MSM))
# equal to the 5-year survival probability, e.g. kaplan-meier at 5 years, i.e. survfit at 5 years:
SF<-survfit(S.MSM~1,data=sel.data)
100*(1-SF$surv[SF$time==max(SF$time[SF$time<=horizon])])
# average probability of MSM
mean(fun.event(lp=lp.MSM, h0=h0.MSM))
```

@@

# manual calculation
lp.score.Rec.manual <- coef.Rec["SNstatus=Positive"]*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)+
              coef.Rec["Age.SN"]*log(data.to.be.imputed[patient.nr,"Age.SN"])+
              coef.Rec["Ulceration=Yes"]*(as.numeric(data.to.be.imputed[patient.nr,"Ulceration"])-1)+
              coef.Rec["Loc_CAT=leg"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="leg")+
              coef.Rec["Loc_CAT=trunk"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="trunk")+
              coef.Rec["Loc_CAT=head & neck"]*as.numeric(data.to.be.imputed[patient.nr,"Loc_CAT"]=="head & neck")+
              coef.Rec["Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)+
              coef.Rec["Rdamcrit"]*log(data.to.be.imputed[patient.nr,"Rdamcrit"])+
              coef.Rec["SNstatus=Positive * Breslow"]*(log(data.to.be.imputed[patient.nr,"Breslow"])-c.Breslow)*(as.numeric(data.to.be.imputed[patient.nr,"SNstatus"])-1)

# lp check
lp.score.Rec.true
lp.score.Rec.try
lp.score.Rec.manual-center.Rec

# probability of recurrence with fun.event
fun.event(lp=lp.score.Rec.try, h0=h0.Rec)
# probability of recurrence manually
1-exp(-h0.Rec*exp(lp.score.Rec.manual-center.Rec))

# check calculation of lp of MSM
lp.score.MSM.true <- predict(f.mi.BS.MSM)[patient.nr]
lp.score.MSM.true
lp.MSM.try <- MSM.cal.fact*(lp.score.Rec.manual-center.Rec)
lp.MSM.try

# probability of MSM with fun.event
fun.event(lp=lp.MSM.try, h0=h0.MSM)
# probability of MSM manually
1-exp(-h0.MSM*exp(MSM.cal.fact*(lp.score.Rec.manual-center.Rec)))
```

### Test with old tool predictions recurrence for negative SN status patients
```{r, eval=TRUE}
rscript.negative <- function(breslow, ulceration, location) {
  S0 <- 0.85837329
  lp <- 0.79351989*log(breslow) +
    0.61912477*I(ulceration == 1) +
    0.29594843*I(location == 1) +
    0.42802046*I(location == 2) +
    0.86053044*I(location == 3) - 0.98860941
  S <- 100 - 100*S0^exp(lp)
  return(S)
}
old.negativeSN.tool <- rscript.negative(breslow=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Breslow"],
        ulceration=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Ulceration"])-1,
        location=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative", "Loc_CAT"])-1)
lp.negative <- predict(f.mi.BS.Rec, newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Negative",], type="lp")
new.negativeSN.tool <- 1-exp(-h0.Rec*exp(lp.negative))
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/new.old.negatives.png"))
plot(x=new.negativeSN.tool*100, y=old.negativeSN.tool,
     xlim=c(0, 100), ylim=c(0, 100),
     xlab="New predictions", ylab="Old predictions",
     main="Predicted recurrence of patients with negative SN status")
abline(a=0, b=1, col="red")
grDevices::dev.off()
```

```{r, eval=TRUE, echo=FALSE}
plot(x=new.negativeSN.tool*100, y=old.negativeSN.tool,
     xlim=c(0, 100), ylim=c(0, 100),
     xlab="New predictions", ylab="Old predictions",
     main="Predicted recurrence of patients with negative SN status")
abline(a=0, b=1, col="red")
```

### Test with old tool predictions recurrence for positive SN status patients
```{r, eval=TRUE}
rscript.positive <- function(age, ulceration, tumor_burden, breslow) {
  S0 <- 0.51762362
  lp <- 0.56109945 * log(age) +
    0.23383770 * log(tumor_burden) +
    0.35489235 * log(breslow) +
    0.36562098 * I(ulceration == 1) - 2.709225
  S <- 100 - 100 * S0^exp(lp)
  return(S)
}
old.positiveSN.tool <- rscript.positive(age=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Age.SN"],
                               ulceration=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Ulceration"])-1,
                               breslow=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Breslow"],
                               tumor_burden=as.numeric(last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive", "Rdamcrit"]))
lp.positive <- predict(f.mi.BS.Rec, newdata=last.imputed.data.set[last.imputed.data.set$SNstatus=="Positive",], type="lp")
new.positiveSN.tool <- 1-exp(-h0.Rec*exp(lp.positive))
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/new.old.positives.png"))
plot(x=new.positiveSN.tool*100, y=old.positiveSN.tool,
     xlim=c(0, 100), ylim=c(0, 100),
     xlab="New predictions", ylab="Old predictions",
     main="Predicted recurrence of patients with positive SN status")
abline(a=0, b=1, col="red")
grDevices::dev.off()
```

```{r, eval=TRUE, echo=FALSE}
plot(x=new.positiveSN.tool*100, y=old.positiveSN.tool,
     xlim=c(0, 100), ylim=c(0, 100),
     xlab="New predictions", ylab="Old predictions",
     main="Predicted recurrence of patients with positive SN status")
abline(a=0, b=1, col="red")
```
